Penalized linear mixed models for structured genetic data

### 2021-03-18 main.tex
Notes from meeting

In many genetic association studies the phenotype of interest is binary, for example the presence or absence of a heritable trait. Unfortunately, the extension of LMMs and penalized LMMs to nonlinear models is not straightforward. As we explain in Section~\ref{sec:methods}, LMMs and penalized LMMs are able to efficiently correct for dependencies among subjects by preprocessing the data by means of a weighted rotation. This procedure does not directly generalize beyond the linear regression case. The situation is certainly not impossible, and many approaches to fitting generalized linear mixed models, or GLMMs, have been proposed in the classical, low-dimensional setting. \anna{However, very little work has been done to extend this to penalized, high-dimensional models. As far we are aware, only \citet{mandt2017sparse} generalize the high-dimensional penalized LMM modeling paradigm to binary phenotypes. They do so by thresholding the outcome variable through a Probit likelihood with additional sparsity constraints and approximate Bayesian inference. However, to the best of our knowledge there is no available software to implement this or any other methods for high-dimensional, penalized methods at this time.} We hope that this situation changes in the near future.

\pb{What is the significance of this Mandt paper? Isn't there a huge literature on GLMMs?}\anna{Mandt is specifically for high-dimensional, penalized LMMs that use the data to infer structure of the errors - not just a regular GLMM. Includes some of the Rakitsch 2013 authors, extends LMM-lasso to binary outcomes in one way.}

@@

All analyses and simulations were performed using R version 3.6.3. The ordinary lasso and PC-lasso methods were implemented using \texttt{glmnet} \citep{glmnet}. As described previously, LMM-lasso-ggmix requires specialized software, but is implemented in the R package \texttt{ggmix} \citep{bhatnagar2020simultaneous}. Standard lasso software such as \texttt{glmnet} can be used to fit the LMM-lasso-Rakitsch method as well, although this requires a fair amount of preprocessing and setting up. To make this process easier, we developed an R package \texttt{penalizedLMM}, available at the github repository \url{https://github.com/areisett/penalizedLMM}, to allow seamless, user-friendly fitting of penalized LMMs based on genotype matrices or PLINK files. \anna{\texttt{penalizedLMM} includes preprocessing functions for genotype imputation using \textbf{imputation rules - how to describe this} where possible, or else the Hardy-Weinberg Equilibrium value.} \texttt{penalizedLMM} additionally implements the elastic net, and the nonconvex penalties, MCP and SCAD.

@@

\begin{figure}[H]
     \centering
     \begin{subfigure}[b]{0.4\textwidth}
         \centering
         \includegraphics[width=\textwidth]{figures/figure_05a.pdf}
         \caption{Precision for each method and number of selected SNPs from the Empirical genome-wide analysis. Single replication case study with $\eta = 0.5, \xi = 0.8$.}
        \label{fig:gw_prc}
     \end{subfigure}
    \hfill
     \begin{subfigure}[b]{0.4\textwidth}
         \centering
         \includegraphics[width=\textwidth]{figures/figure_05b.pdf}
         \caption{MSE for each method and number of selected SNPs from the Empirical genome-wide analysis. Single replication case study with $\eta = 0.5, \xi = 0.8$.}
        \label{fig:gw_mse}
     \end{subfigure}
     \begin{subfigure}[b]{0.15\textwidth}
         \centering
         \includegraphics[width=\textwidth]{figures/figure_05c.pdf}
     \end{subfigure}
\end{figure}

>>>>

In many genetic association studies the phenotype of interest is binary, for example the presence or absence of a heritable trait. Unfortunately, the extension of LMMs and penalized LMMs to nonlinear models is not straightforward. As we explain in Section~\ref{sec:methods}, LMMs and penalized LMMs are able to efficiently correct for dependencies among subjects by preprocessing the data by means of a weighted rotation. This procedure does not directly generalize beyond the linear regression case.

\anna{other approaches are possible, for example \citet{mandt2017sparse} generalize the penalized lmm modeling paradigm to binary phenotypes by thresholding the outcome variable through a probit likelihood. However, there is no work that we are aware of that explicitly addresses the use of such models for GWAS.}

\anna{idea: not impossible, important,b ut not enough has been done for us to include it in the review}

The situation is certainly not impossible, and many approaches to fitting generalized linear mixed models, or GLMMs, have been proposed in the classical, low-dimensional setting. \anna{However, very little work has been done to extend this to penalized, high-dimensional models. As far we are aware, only \citet{mandt2017sparse} generalize the high-dimensional penalized LMM modeling paradigm to binary phenotypes. They do so by thresholding the outcome variable through a Probit likelihood with additional sparsity constraints and approximate Bayesian inference. However, to the best of our knowledge there is no available software to implement this or any other methods for high-dimensional, penalized methods at this time.} We hope that this situation changes in the near future.

@@

All analyses and simulations were performed using R version 3.6.3. The ordinary lasso and PC-lasso methods were implemented using \texttt{glmnet} \citep{glmnet}. As described previously, LMM-lasso-ggmix requires specialized software, but is implemented in the R package \texttt{ggmix} \citep{bhatnagar2020simultaneous}. Standard lasso software such as \texttt{glmnet} can be used to fit the LMM-lasso-Rakitsch method as well, although this requires a fair amount of preprocessing and setting up. To make this process easier, we developed an R package \texttt{penalizedLMM}, available at the github repository \url{https://github.com/areisett/penalizedLMM}, to allow seamless, user-friendly fitting of penalized LMMs based on genotype matrices or PLINK files. \anna{\texttt{penalizedLMM} includes preprocessing functions for genotype imputation using functions for imputation based on linkage disequilibrium with tag SNPs where possible, or else the Hardy-Weinberg Equilibrium value.} \texttt{penalizedLMM} additionally implements the elastic net, and the nonconvex penalties, MCP and SCAD.

\anna{functions for imputation based on linkage disequilibrium with tag SNPs.}

@@

\anna{Make sure this gets an overall figure number}
\begin{figure}[H]
     \centering
     \begin{subfigure}[b]{0.4\textwidth}
         \centering
         \includegraphics[width=\textwidth]{figures/figure_05a.pdf}
         \caption{Precision for each method and number of selected SNPs from the Empirical genome-wide analysis. Single replication case study with $\eta = 0.5, \xi = 0.8$.}
        \label{fig:gw_prc}
     \end{subfigure}
    \hfill
     \begin{subfigure}[b]{0.4\textwidth}
         \centering
         \includegraphics[width=\textwidth]{figures/figure_05b.pdf}
         \caption{MSE for each method and number of selected SNPs from the Empirical genome-wide analysis. Single replication case study with $\eta = 0.5, \xi = 0.8$.}
        \label{fig:gw_mse}
     \end{subfigure}
     \begin{subfigure}[b]{0.15\textwidth}
         \centering
         \includegraphics[width=\textwidth]{figures/figure_05c.pdf}
     \end{subfigure}
\end{figure}
